---
title: "Masting"
author: "Gleb Kozienko"
date: "2023-06-01"
output: pdf_document
---

This document has functions and code chunks I was using to get climate data and add it to our masting data.




1) Get needed libraries 


a) essential libraries
```{r}
library(readr)
library(dplyr)
library(stringr)   
library(tidyverse)
library(lubridate)
library(ggplot2)
```

b) libraries for climate data acquisition 
```{r}
library(devtools)#needed for using install_github()
install_github("rspatial/raster")
library(sp)
library(raster)

```

c) other libraries
```{r}
library(tidygeocoder)#needed for reverse_geo() -- for address

```




2) Get dataset

Here, MASTREE+_selection_2023-06-01.csv comes from https://mastreeplus.shinyapps.io/mastreeplus/.

Instructions:
1) click "Select all" in the top right corner tab named "Filters".
2) click "Download selection" located under the map of the world in the main tab named "Map / Table"

```{r}
#here you should read in the .csv dataset you downloaded from the link above
#dont forget to modify the folder path
MASTREE <- read_csv("Desktop/CURI/R code/Mastree datasets/MASTREE+_selection_2023-06-01.csv") 


```





FOR FUTURE MASTING RESEARCHERS:
Copernicus website: https://cds.climate.copernicus.eu/#!/home

Dataset that I am using: https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-land-monthly-means?tab=overview

Climatic data that we get from Copernicus website starts from 1950. Thus, I filter out all the records from MASTREE dataset that are dated before 1950. 
Result of this filtering is "MASTREE_W_ADDRESS_years_of_climate" from the chunk below.

Hopefully, you have access to folder named "Various mastree datasets". In this folder you can find "MASTREE_W_ADDRESS_years_of_climate.csv". If, for some reason, you do not have access to this CSV, feel free to run chunks below top get this data.
```{r}
MASTREE_W_ADDRESS_years_of_climate <- read_csv("Desktop/CURI/R code/Mastree datasets/MASTREE_W_ADDRESS_years_of_climate.csv")
```




STEPS TO GET MASTREE_W_ADDRESS_years_of_climate IF CSV IS MISSING:


1) Add address
```{r}
MASTREE_W_ADDRESS<-MASTREE|>
  mutate(address=reverse_geo(Latitude, Longitude)[[3]])
```


2) Update coords
```{r}
#round latitude and lon gitude to the first decimal. I am doing it because climate data from Copernicus website (keep reading to learn more about it) has resolution of 0.1 degree. 
MASTREE_W_ADDRESS<-MASTREE_W_ADDRESS|>
  mutate(Lat_adj=round(Latitude, digits = 1),
         Long_adj=round(Longitude, digits = 1),
         )

#create new column, that has coords rounded to the first decimal
MASTREE_W_ADDRESS<-MASTREE_W_ADDRESS|>
  mutate(Coords=paste(Lat_adj, Long_adj, sep = ", "))
```

3)  filter out all the records from MASTREE dataset that are dated before 1950. 
```{r}
 MASTREE_W_ADDRESS_years_of_climate<-MASTREE_W_ADDRESS|>
   filter(Start %in% c(1950:2021))
```





TO GET THE CLIMATE DATA FROM COPERNICUS, I AM USING THE FUNCTIONS BELOW.
THESE FUNCTIONS ARE MODIFICATIONS OF THE FUNCTIONS PRESENTED IN "ForeMast" R package. 


Copernicus website: https://cds.climate.copernicus.eu/#!/home

Dataset that I am using: https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-land-monthly-means?tab=overview

foreMast: an R package for predicting beech (Fagus sylvatica L.) masting events in European countries. 

Link to the scientific paper that describes "ForeMast" R package
https://annforsci.biomedcentral.com/articles/10.1007/s13595-021-01109-5

Link to foremast github:
https://github.com/uchiavetta/foreMast

Link to github folder that contains functions i have modified:
https://github.com/uchiavetta/foreMast/tree/master/R



Define function that downloads climate data in .nc format and saves it at specified directory. 

Note: in the chunk below you can see that climatic variables to be downloaded are specified by the object called "variable" that is a string that contains names of variables from "reanalysis-era5-land-monthly-means" dataset. For descriptions of variables used (or if you need to get other variables), see the following website: https://confluence.ecmwf.int/display/CKB/ERA5-Land%3A+data+documentation
or
https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-land-monthly-means?tab=overview
```{r}
# Function to download ERA5-Land climate data from the Copernicus website using the ecmwfr package.

# Arguments:
# U_ID: Your CDS user ID.
# API_Key: Your CDS API key.
# lat: Latitude of the location for which data is to be downloaded.
# lon: Longitude of the location for which data is to be downloaded.
# sPath: Directory path where the downloaded files will be saved. (Default is the current working directory).
# site_id: Identifier for the location. For this argument you need to supply coordinate string in the form latitude_longitude (example: "-0.2_11.6", or "48.5_9.1", or "48.5_-9.1")

cdsDownload <- function(U_ID, API_Key, lat, lon, sPath = getwd(), site_id = " ") {

  cat("\n")
  cat("Check the download request progress at https://cds.climate.copernicus.eu/cdsapp#!/yourrequests")
  cat("\n")
  cat("\n")

  # Create the target directory for saving the downloaded files, suppressing warnings if the directory already exists.
  dir.create(sPath, showWarnings = FALSE)

  # Set the CDS API key using the ecmwfr package.
  ecmwfr::wf_set_key(user = U_ID, key = API_Key, service = 'cds')

  # Define the start and end years for the data download.
  start.year = 1950
  end.year = format(Sys.time(), "%Y") #Current year

  # Create a character vector of years from start.year to the current year.
  yearL = as.character(start.year:end.year)

  # Create a character vector of months, zero-padded with leading zeros (01 to 12).
  monthL = stringr::str_pad(1:12, 2, pad = "0")

  # Define the data request parameters as a list.
  request <- list(
    format = "netcdf", # Requested data format (netCDF).
    product_type = "monthly_averaged_reanalysis", # Type of product (monthly reanalysis data).
    
     # Variables of interest to be downloaded. 
    #you can see the names of the variables in the description of "reanalysis-era5-land-monthly-means" dataset here: https://confluence.ecmwf.int/display/CKB/ERA5-Land%3A+data+documentation
    variable = c("2m_temperature",
                 "total_precipitation",
                 "soil_temperature_level_3",
                 "soil_temperature_level_4",
                 "surface_solar_radiation_downwards",
                 "volumetric_soil_water_layer_3",
                 "volumetric_soil_water_layer_4",
                 "leaf_area_index_low_vegetation",
                 "leaf_area_index_high_vegetation"),
    year = yearL, # Years for which data is requested.
    month = monthL, # Months for which data is requested.
    time = "00:00", # Time of the day for the data (00:00).
    area = c(lat, lon, lat, lon), # Geographic bounding box (latitude and longitude).
    dataset_short_name = "reanalysis-era5-land-monthly-means", # Short name of the dataset.
    target = paste0("ClimData_", site_id, ".nc") # File name for the downloaded data.
  )

  # Perform the data request using the ecmwfr package and save the data as a netCDF file.
  file <- ecmwfr::wf_request(
    user = U_ID,
    request = request,
    transfer = TRUE,
    path = sPath
  )
}
```


Define function that reads in climate data, turns it into data frame
```{r}



mastFaSyl <- function(fName, csv.coordinates = c(NULL, NULL), weighting = "", weights = c(NULL, NULL), adverseEvent = "N") {
  
  
  distanceFromPoint <- NULL  # Initialize distanceFromPoint variable
  
  
  
  nc <- ncdf4::nc_open(fName)  # Open the netCDF file with climate data
  data.years <- ncdf4::ncvar_get(nc, "time")  # Get the time variable from the netCDF file
  lat <- base::round(ncdf4::ncvar_get(nc, "latitude"), 1)  # Get the latitude variable from the netCDF file and round it to one decimal place
  lon <- base::round(ncdf4::ncvar_get(nc, "longitude"), 1)  # Get the longitude variable from the netCDF file and round it to one decimal place
  
  Coords=paste(lat, lon, sep = ", ") #get a string of coordinates in the form "latitude, longitude"
  
  
  # This function extracts the time from the nc file, converting it from seconds-from-1900 to year
  conversion <- function(x) {
    converted <- base::floor((x/24)/365)  # Convert seconds to days, then to years
    converted <- converted + 1900  # Add 1900 to get the actual year
    return(converted)
  }
  
  
  
  data.year <- base::lapply(data.years, conversion)  # Apply the conversion function to each element of data.years
  start.year <- data.year[[1]]  # Get the start year from the first element of data.year
  
  
  
  # Time series of the climate variables
  
  #in the lines below, "varid" represents climatic variables of interest. 
  #see website https://confluence.ecmwf.int/display/CKB/ERA5-Land%3A+data+documentation
  #on this website, in the table that describes variables, find "shortName" column. 
  #notice that i assign names from "shortName" column to the variable "varid"
  
  t2m <- stats::ts(ncdf4::ncvar_get(nc, varid = "t2m") - 273.15, frequency = 12,
                   start = c(as.numeric(start.year), 1))  # Get the temperature variable and convert it from Kelvin to Celsius
  tp <- stats::ts(ncdf4::ncvar_get(nc, varid = "tp"), frequency = 12)  # Get the precipitation variable
  
  stl3 <- stats::ts(ncdf4::ncvar_get(nc, varid = "stl3") - 273.15, frequency = 12)  
  
  stl4 <- stats::ts(ncdf4::ncvar_get(nc, varid = "stl4") - 273.15, frequency = 12)

  ssrd <- stats::ts(ncdf4::ncvar_get(nc, varid = "ssrd"), frequency = 12)

  swvl3 <- stats::ts(ncdf4::ncvar_get(nc, varid = "swvl3"), frequency = 12)

  swvl4 <- stats::ts(ncdf4::ncvar_get(nc, varid = "swvl4"), frequency = 12)

  lai_lv <- stats::ts(ncdf4::ncvar_get(nc, varid = "lai_lv"), frequency = 12)

  lai_hv <- stats::ts(ncdf4::ncvar_get(nc, varid = "lai_hv"), frequency = 12)

  
  
  
  
  t2m.df <- data.frame(stats::.preformat.ts(t2m), stringsAsFactors = FALSE)  # Convert temperature time series to data frame
  tp.df <- data.frame(stats::.preformat.ts(tp), stringsAsFactors = FALSE)  # Convert precipitation time series to data frame
  
  stl3.df <- data.frame(stats::.preformat.ts(stl3), stringsAsFactors = FALSE)

  stl4.df <- data.frame(stats::.preformat.ts(stl4), stringsAsFactors = FALSE)

  ssrd.df <- data.frame(stats::.preformat.ts(ssrd), stringsAsFactors = FALSE)

  swvl3.df <- data.frame(stats::.preformat.ts(swvl3), stringsAsFactors = FALSE)

  swvl4.df <- data.frame(stats::.preformat.ts(swvl4), stringsAsFactors = FALSE)

  lai_lv.df <- data.frame(stats::.preformat.ts(lai_lv), stringsAsFactors = FALSE)

  lai_hv.df <- data.frame(stats::.preformat.ts(lai_hv), stringsAsFactors = FALSE)

  
  
  year <- c()  # Initialize an empty vector for year
  month <- c()  # Initialize an empty vector for month
  temp_2m <- c()  # Initialize an empty vector for average temperature
  tot_precip <- c()  # Initialize an empty vector for average precipitation
  soil_temp_l3 <- c()
  soil_temp_l4 <- c()
  surf_sol_rad_down <- c()
  vol_soil_water_l3 <- c()
  vol_soil_water_l4 <- c()
  leaf_area_ind_lv <- c()
  leaf_area_ind_hv <- c()
  
  
  
  for (i in 1:length(data.year)) {
    t2m.s <- t2m.df[1, i]  # get the average temperature for the month
    tp.s <- tp.df[1, i]  # get the average precipitation for the month
    stl3.s <-  stl3.df[1, i]
    stl4.s <- stl4.df[1, i]
    ssrd.s <- ssrd.df[1, i]
    swvl3.s <- swvl3.df[1, i]
    swvl4.s <- swvl4.df[1, i]
    lai_lv.s <- lai_lv.df[1, i]
    lai_hv.s <- lai_hv.df[1, i]
    
    
    
    
    year <- c(year, data.year[[i]])  # Append the year to the year vector
    month <- c(month, (i - 1) %% 12 + 1)  # Append the month to the month vector
    temp_2m <- c(temp_2m, t2m.s)  # Append the average temperature to the avg_temp vector
    tot_precip <- c(tot_precip, tp.s)  # Append the average precipitation to the avg_precip vector
    soil_temp_l3 <- c(soil_temp_l3, stl3.s)
    soil_temp_l4 <- c(soil_temp_l4, stl4.s)
    surf_sol_rad_down <- c(surf_sol_rad_down, ssrd.s)
    vol_soil_water_l3 <- c(vol_soil_water_l3, swvl3.s)
    vol_soil_water_l4 <- c(vol_soil_water_l4, swvl4.s)
    leaf_area_ind_lv <- c(leaf_area_ind_lv, lai_lv.s)
    leaf_area_ind_hv <- c(leaf_area_ind_hv, lai_hv.s)
  }
  
  
   
  Coord_vec <- rep(Coords, length(temp_2m))
  
  result <- data.frame(Coord_vec,
                       year, 
                       month,
                       temp_2m,
                       tot_precip,
                       soil_temp_l3,
                       soil_temp_l4,
                       surf_sol_rad_down,
                       vol_soil_water_l3,
                       vol_soil_water_l4,
                       leaf_area_ind_lv,
                       leaf_area_ind_hv
                      )  # Create a data frame with the year, month, average temperature, and average precipitation
  
  return(result)  # Return the resulting data frame
}

```










GET CLIMATE VARIABLES FOR EACH COORDINATE 

FOR FUTURE MASTING RESEARCHERS: Feel free not to use this code unless you need some extra data. Hopefully, you have access to "MASTREE_climate_2.csv" dataset that should be located in "Various mastree datasets" folder. MASTREE_climate_2.csv contains about a half of original MASTREE dataset together with corresponding climate variables. 

However, here is how to obtain climate variables. 

1) start by creating a list of coordinates arranged from most observations per coordinate to the least. 
```{r}

Coords_by_obs<-MASTREE_W_ADDRESS_years_of_climate|>
  group_by(Coords)|>
  summarize(count=n())|>
  arrange(desc(count))|>
  print(n=40)
```

2) Get Copernicus API info

a) Go to https://cds.climate.copernicus.eu/#!/home
b) register an account
c) After you logged into your account, you should see the red button "Logout" in the top right corner of the website. Immediately on the left from this button, you will see your username. Click on your username. 
d) after you clicked on your username, you will see page with the text that says: "User profile : <your username>". Below this text there will be two tabs: "View" and "Edit". 
Stay on the "View" tab and scroll to the very bottom to bind a "API key" title. Under this title you will see UID number (it's your user_id) and API Key number (it's your user_key)
```{r}
 user_id <- "123456" #example, not working user id
 user_key <- "58c80400-0cba-4a89-qwe1-e6def123f90b" #example, not working key
 dir="~/Desktop/CURI/R code" #don't forget to change it to the directory you desire
```


3) Specify the coordinates of interest and create additional columns for coordinates in different formats.
I will use coordinates of different formats as a part of file naming.
Some symbols as "," and "-" cannot serve as a file names, thus we need different formats.

```{r}

#while coords_by_obs have all the coordinates from MASTREE_W_ADDRESS_years_of_climate, coords_set contains only the coordinates for which you want to get the climate data

Coords_set<-Coords_by_obs[c(1:10),1]

#coords_set contains only the coordinates for which you want to get the climate data!!!!!!

Coords_set[c('Lat', 'Lon')] <- str_split_fixed(Coords_set$Coords, ', ', 2)

Coords_set<-Coords_set|>
  mutate(Coords2=paste(Lat, Lon, sep = "_"))


Coords_set<-Coords_set|>
  mutate(Coords3=str_replace_all(Coords2,"-", "minus_"))

```


4) This code downloads separate climate .nc files for each coordinate in Coords_set
```{r}

for  (n in 1:length(Coords_set$Coords)){ #iterate through each coordinate
  
  N=as.numeric(Coords_set$Lat[n]) #get latitude
  W=as.numeric(Coords_set$Lon[n])#get longitude
  id=(Coords_set$Coords2[n]) #id sets the unique file name. see the function cdcDownload to learn more about file naming system
  
  cdsDownload(U_ID = user_id, API_Key = user_key, lat = N, lon = W, sPath = dir, site_id = id)
  
}

```

5) This code accesses separate climate .nc files for each coordinate in Coords_set, unpacks it, adds this data to MASTREE data at that coordinate, then saves MASTREE with climate at particular coordinate as a .csv file
```{r}
for  (n in 1:length(Coords_set$Coords)){
  
coords_for_filter <- Coords_set$Coords[n] #this coordinate will be used to filter MASTREE dataset
  
coords_for_directory_nc <- Coords_set$Coords2[n] #this coordinate will be used to access .nc file

coords_for_directory_csv <- Coords_set$Coords3[n]#this coordinate will be used 
  

#specify the directory into which you ALREADY saved the .nc files:
#"~/Desktop/CURI/R code/Raw Climate Data/ClimData_"  ---change this to the direction of the filder with .nc files
directory_nc <- paste("~/Desktop/CURI/R code/Raw Climate Data/ClimData_", coords_for_directory_nc, ".nc", sep = "") 
  

#specify the directory into which you WILL BE saving the .csv files:
#"Desktop/CURI/R code/Climate CSVs/MASTREE_"  ---- change this to the direction of the filder with .nc files

directory_csv <-  paste("Desktop/CURI/R code/Climate CSVs/MASTREE_", coords_for_directory_csv, "_climate.csv", sep = "") 
  

#this function unpacks .nc file for particular coord
Data_at_coord<-mastFaSyl(directory_nc, weighting = standard)

#this code acquires MASTREE data for particular coord
MASTREE_at_coord <- MASTREE_W_ADDRESS_years_of_climate|>
  filter(Coords == coords_for_filter)
  
#this code adds climate to the MASTREE data for particular coord
MASTREE_at_coord_climate <- MASTREE_at_coord|>
  left_join(Data_at_coord,by=c("Year"="year"), relationship = "many-to-many")
  

#this function saves MASTREE and Climate data for particular coordinate as a .csv file
write_csv(MASTREE_at_coord_climate, file = directory_csv)
  
}
```



6) This code is similar to the code above. It also accesses separate climate .nc files for each coordinate in Coords_set, unpacks it, adds this data to MASTREE data at that coordinate, then saves MASTREE with climate at particular coordinate as a .csv file. 
But, in addition to the code above, it joins MASTREE and climate data at each coordinate into one big dataset.

```{r}

coords_for_filter <- Coords_set$Coords[n] #this coordinate will be used to filter MASTREE dataset
  
coords_for_directory_nc <- Coords_set$Coords2[n] #this coordinate will be used to access .nc file

  
#specify the directory into which you ALREADY saved the .nc files:
#"~/Desktop/CURI/R code/Raw Climate Data/ClimData_"  ---change this to the direction of the filder with .nc files
directory_nc <- paste("~/Desktop/CURI/R code/Raw Climate Data/ClimData_", coords_for_directory_nc, ".nc", sep = "") 


#this function unpacks .nc file for particular coord
Data_at_coord<-mastFaSyl(directory_nc, weighting = standard)

#this code acquires MASTREE data for particular coord
MASTREE_at_coord <- MASTREE_W_ADDRESS_years_of_climate|>
  filter(Coords == coords_for_filter)
  
#this code adds climate to the MASTREE data for particular coord
MASTREE_at_coord_climate <- MASTREE_at_coord|>
  left_join(Data_at_coord,by=c("Year"="year"), relationship = "many-to-many")



MASTREE_climate_EXTRA_1 <- MASTREE_at_coord_climate

for  (n in 2:length(Coords_set$Coords)){
  
coords_for_filter <- Coords_set$Coords[n] #this coordinate will be used to filter MASTREE dataset
  
coords_for_directory_nc <- Coords_set$Coords2[n] #this coordinate will be used to access .nc file
  

#specify the directory into which you ALREADY saved the .nc files:
#"~/Desktop/CURI/R code/Raw Climate Data/ClimData_"  ---change this to the direction of the filder with .nc files
directory_nc <- paste("~/Desktop/CURI/R code/Raw Climate Data/ClimData_", coords_for_directory_nc, ".nc", sep = "") 
  

#this function unpacks .nc file for particular coord
Data_at_coord<-mastFaSyl(directory_nc, weighting = standard)

#this code acquires MASTREE data for particular coord
MASTREE_at_coord <- MASTREE_W_ADDRESS_years_of_climate|>
  filter(Coords == coords_for_filter)
  
#this code adds climate to the MASTREE data for particular coord
MASTREE_at_coord_climate <- MASTREE_at_coord|>
  left_join(Data_at_coord,by=c("Year"="year"), relationship = "many-to-many")
  
#join data for different coordinates together
MASTREE_climate_EXTRA_1 <- rbind(MASTREE_climate_EXTRA_1, MASTREE_at_coord_climate)
  
  
}
```






